Pipit: Reactive Systems in \fstar ~(Extended Abstract)

### 2023-05-03 2023pipit/Main.tex
initial writeup

Cite: Kind2, Lustre, Vélus.
Antilock brakes paper.
The original Kind~\cite{hagen2008verifying}.

\begin{verbatim}
  let veh_speed_estimate ()
\end{verbatim}

@@

The grammar of Pipit is defined in \autoref{f:core-grammar}.
The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and applications ($e~e'$); however, it does not include any form for defining functions other than reusing closed functions from the \fstar{} meta-language.
Most of the expression forms are equivalent to a simplified, clock-free primitives from standard Lustre-style languages \CITE.
The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.
Streams can be composed together using the \emph{then} notation ($\xthen{e}{e'}$) which denotes that the value of stream $e$ is used for the very first step, followed by the latest values from stream $e'$ for subsequent steps.

@@

The semantics here, particularly the recursive rule, differ from previous semantics of Lustre.
In the reactive semantics of \CITEME{88}, as well as the coinductive semantics of Vélus \CITEME{Vélus}, the semantics for mutually recursive definitions is locally non-deterministic, in that the semantics does not define which particular order mutual definitions are evaluated in.
This local non-determinism complicates certain proofs, even though the overall evaluation \emph{is} deterministic \TODO{vague}.

In contrast, the semantics presented here is syntax-directed and completely deterministic.
We have a mechanised proof of variant of progress theorem for well-formed (\emph{causal}) programs as well as closure under substitution and determinism, which is described in \REF.
However, the substitution-based semantics here can duplicate expressions, and encoding mutually recursive definitions requires multiple substitutions.
For this reason, the semantics here is not suitable for actual execution.

>>>>

We are interested in verifying an antilock braking system for a motorcycle\cite{huang2010design}.

\begin{tabbing}
  @MM@\= @let @\= \kill
  @let@ veh\_speed\_estimate $\omega_F$ $\omega_R$ $a_z$ = \\
    \> @let@ $v_F$ = $\omega_F \cdot $ circumference @in@ \\
    \> @let@ $v_R$ = $\omega_R \cdot $ circumference @in@ \\
    \\
    \> @let rec@ $\floor{\hat{v}}$ = @if@ $v_F \approx_\epsilon v_R$ \\
    \> \> @then@ $\min v_F v_R$ \\
    \> \> @else@ $(\xthen{\min v_F v_R}{\xfby{0}{\floor{\hat{v}}}}) + a_z - \epsilon$ @in@ \\
    \> @let rec@ $\ceil{\hat{v}}$ = @if@ $v_F \approx_\epsilon v_R$ \\
    \> \> @then@ $\max v_F v_R$ \\
    \> \> @else@ $(\xthen{\max v_F v_R}{\xfby{0}{\ceil{\hat{v}}}}) + a_z + \epsilon$ @in@
    \\
    \> $(\floor{\hat{v}}, \ceil{\hat{v}})$
\end{tabbing}

Cite: Kind2\cite{champion2016kind2}, Lustre\cite{caspi1995functional}, Vélus\cite{bourke2017formally}.
The original Kind\cite{hagen2008scaling}.

Past-time modal temporal logic \cite{lichtenstein1985glory}
Lustre support for temporal logic \cite{halbwachs1993executable}

@@

The grammar of Pipit is defined in \autoref{f:core-grammar}.
The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and applications ($e~e'$); however, it does not include any form for defining functions other than reusing closed functions from the \fstar{} meta-language.
Most of the expression forms are equivalent to a simplified, clock-free primitives from standard Lustre-style languages \cite{caspi1995functional}.
The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.
Streams can be composed together using the \emph{then} notation ($\xthen{e}{e'}$) which denotes that the value of stream $e$ is used for the very first step, followed by the latest values from stream $e'$ for subsequent steps.

@@

The semantics here, particularly the recursive rule, differ from previous semantics of Lustre.
In the reactive semantics of \cite{caspi1995functional}, as well as the coinductive semantics of Vélus \CITE{bourke2017formally find best paper}, the semantics for mutually recursive definitions is locally non-deterministic, in that the semantics does not define which particular order mutual definitions are evaluated in.
This local non-determinism complicates certain proofs, even though the overall evaluation \emph{is} deterministic \TODO{vague}.

In contrast, the semantics presented here is syntax-directed and completely deterministic.
We have a mechanised proof of variant of progress theorem for well-formed (\emph{causal}) programs as well as closure under substitution and determinism, which is described in \autoref{ss:meta-theory}.
However, the substitution-based semantics here can duplicate expressions, and encoding mutually recursive definitions requires multiple substitutions.
For this reason, the semantics here is not suitable for actual execution.

\subsection{Causality}
\label{ss:causality}

Talk about the causality restriction.
Does expression $e$ have a dependency on the most-recent value of expression $e'$?

\begin{code}
let rec direct_dependency (e e': exp) : bool =
  if e = e' then true
  else
  match e with
  | XVal _ -> false
  | XVar x' -> false
  | XPrim2 _ e1 e2 -> direct_dependency e1 e' || direct_dependency e2 e'
  | XIte ep et ef -> direct_dependency ep e' || direct_dependency et e' || direct_dependency ef e'
  | XThen e1 e2 -> direct_dependency e1 e' || direct_dependency e2 e'
  | XPre _ -> false
  | XMu e1 -> direct_dependency e1 (lift e' 0)
  | XLet e1 e2 -> direct_dependency e1 e' || direct_dependency e2 (lift e' 0)

let rec causal (e: exp): bool =
  match e with
  | XVal _ -> true
  | XVar x' -> true
  | XPrim2 _ e1 e2 -> causal e1 && causal e2
  | XIte ep e1 e2 -> causal ep && causal e1 && causal e2
  | XThen e1 e2 -> causal e1 && causal e2
  | XPre e1 -> causal e1
  | XMu e1 -> causal e1 && not (direct_dependency e1 (XVar 0))
  | XLet e1 e2 -> causal e1 && causal e2
\end{code}

\begin{lemma}[no dependency, no difference]
If $e$ does not have a direct depend on $x$, then any evaluation of $e$ cannot depend on the current value of $x$.

For all expressions $e$ and well-scoped variables $x$,
if $\neg \mbox{directly} ~e~ x$ and
$\bigstep{\Sigma; (\sigma, x \mapsto v_x)}{e}{v}$
then for any $v_x'$, evaluation is unchanged
$\bigstep{\Sigma; (\sigma, x \mapsto v_x')}{e}{v}$
\end{lemma}

We also need:
\begin{lemma}[Bigstep recursive XMu]
  If a recursive expression $\xrec{x}{e}$ evaluates to some sequence of values $v$, then evaluating the recursive expression doesn't depend on the last value in the sequence:

  if $\bigsteps{\Sigma}{\xrec{x}{e}}{v_0 v_1 \cdots v_{n-1} v_n }$
  then for any $v'$,
  $\bigsteps{\Sigma \triangleright^{\uparrow} (v_0 v_1 \cdots v_{n-1} v')}{e}{v_0 v_1 \cdots v_{n-1} v_n}$
\end{lemma}

% let bigstep_recursive_XMu (#outer #inner: nat) (e: exp)
%   (streams: C.table (outer + 1) inner)
%   (vs: C.vector value outer) (v v': value)
%   (hBSmu: bigstep streams (XMu e) (v :: vs)):
%     bigstep (C.table_map_append (C.table_of_values (v' :: vs)) streams) e (v :: vs) =

\begin{lemma}[progress]
  if expression is causal, then for any valid stream history it should evaluate to a value:

  if $\typing{\Gamma}{e}{\tau}$ and $e \mbox{causal}$ and $\Gamma \vdash* \Sigma$, then exists $v$ such that
  $\bigstep{\Sigma}{e}{v}$ and $\mtyping{}{v}{\tau}$
\end{lemma}

\begin{lemma}[monotonic progress]
  if an expression $e$ evaluates to a sequence $vs$ under a stream history $\Sigma$ and a new stream store $\sigma$ is valid, then
  there exists some $v'$ such that
  $\bigsteps{\Sigma; \sigma}{e}{vs; v}$
\end{lemma}

\subsection{Meta-theory}
\label{ss:meta-theory}

We use a typed representation, which ensures an intrinsic variant of preservation.
For programs that satisfy the causality restriction, we have proved a variant of the progress theorem.

\begin{code}
let bigstep_deterministic
  (#streams: list (C.row 'c))
  (#e: exp 'c 'a)
  (#v1 #v2: 'a)
  (hBS1: bigstep streams e v1) (hBS2: bigstep streams e v2):
    Lemma (ensures (v1 == v2)) (decreases hBS1) =
  bigstep_proof_equivalence hBS1 hBS2
\end{code}

Monotonicity

Substitution-closure
\begin{code}
let rec bigstep_substitute_as_var (e p: exp)
  (streams1: C.table outer inner1)
  (streams2: C.table outer inner2)
  (vsp vsep: C.vector value outer)
  (hBSp: bigstep (C.table_map_append streams1 streams2) p vsp)
  (hBSep: bigstep (C.table_map_append streams1 streams2) (subst e inner1 p) vsep):
    Tot (bigstep (C.table_map_append streams1 (C.table_map_append (C.table_of_values vsp) streams2)) e vsep) (decreases hBSep) =

\end{code}

\section{Proving it with transition systems}

Describe the translation from Pipit expressions to labelled transition systems.

K-induction

Statement of LTS soundness:
\begin{code}
let system_eval_complete'
    (#outer: nat) (#vars: nat)
    (e: exp { causal e /\ wf e vars })
    (streams: C.table outer vars)
    (vs: C.vector value outer)
    (hBS: bigstep streams e vs):
      Lemma (exists (s': option (state_of_exp e)).
        system_stepn'
          (system_of_exp e vars)
          (C.Table?._0 streams) vs s') =
    match outer with
    | 0 -> assert
      (system_stepn'
        (system_of_exp e vars)
        (C.Table?._0 streams) vs None)
    | _ -> system_eval_complete e streams vs hBS
\end{code}

Tactics: \cite{martinez2019meta}.

Invariant generation: ICE\cite{garg2014ice}; template-based\cite{kahsai2011instantiation}; PDR explanation\cite{een2011efficient}; IC3 first introduced\cite{bradley2011sat}

\section{Execution}

We have implemented an unverified imperative code generation transform, by making use of \fstar{}'s existing C code extraction support\cite{protzenko2017verified}.
% This transform generates \fstar{} functions in the

This allows us to generate C code which executes in bounded space on embedded devices.

Other compilers for Lustre-style languages, such as Vélus\cite{bourke2017formally} and Heptagon \cite{gerard2012modular} ensure that the generated code can be executed in bounded memory and that each step takes a bounded time.
With a carefully chosen set of primitives, the core streaming operations shown here would ensure bounded memory and time.
For practicality, however, we currently allow embedding arbitrary total functions from the \fstar{} meta-language, which are not necessarily bounded in time.
(Is this the same as Lucid Synchrone and Zélus? \CITE)

\section{Related work}

hardware language in Agda \cite{harrison2021mechanized}

Refinement types for Zélus \cite{chen2022synchronous}

\section{Conclusion and future work}

Future work: verify imperative code generation; verify CSE; case studies: antilock braking; clocks \cite{caspi1995functional}; contracts \cite{champion2016kind2}.
